Smooth risk-based predictive approaches to treatment effect heterogeneity: A simulation study

### 2021-06-15 submission/manuscript.rmd
fixing typos

---
title: "Smooth risk-based predictive approaches to treatment effect heterogeneity: A simulation study"
output:
    bookdown::pdf_document2: default
    bookdown::word_document2:
        reference_docx: reference.docx
geometry: margin=1.0in
toc: false
font-size: 11pt
header-includes:
  - \renewcommand*\familydefault{\sfdefault}
  - \usepackage{setspace}
  - \usepackage{amsmath}
  - \doublespacing
  - \usepackage[left]{lineno}
editor_options: 
  chunk_output_type: console
bibliography: references.bib
csl: jce.csl
---

@@

In the simulated datasets of the base-case scenario treatment was allocated at
random using a 50/50 split. For each patient we simulated $8$ baseline
covariates, where $x_1,\dots,x_4\sim N(0, 1)$ and $x_5,\dots,x_8\sim B(1, 0.2)$.
Outcomes for patients in the control arm were generated from a logistic
regression model including all baseline covariates. Coefficient values were
such, so that the prediction model had an AUC of $0.75$ and an event rate of
$20\%$ in the control arm was achieved. Outcomes in the treatment arm were
created using the same logistic regression model, including a constant treatment
effect odds ratio (OR) of $0.8$. The generated samples of the base-case scenario
were of size $n=4,250$ ($80\%$ power for the detection of an unadjusted OR of
$0.8$).

@@

A true logistic regression model with a constant treatment effect (constant OR)
implies that outcome risk in the treatment arm is a straight line parallel to
the first diagonal on the *log-odds* scale, with distance equal to $\log(\text{OR})$. 
We assessed the effect of stronger and absent relative treatment effects
($\text{OR}=0.5$ or $\text{OR}=1$). We also relaxed the assumption that the
line should be parallel to the diagonal, considering moderate and stronger
linear deviations. Finally, we dropped the assumption of linearity allowing for
quadratic deviations.

@@

The **stratified HTE method** was suggested as an alternative to traditional
subgroup analyses. Patients are stratified into equally-sized risk strata---in
this case based on risk quartiles. Absolute effects are estimated using the
differences in event rates between treatments within risk quarters. We
considered this approach as a reference, expecting it to perform worse than the
other candidates, as its objective is not individual benefit prediction.

We also considered a set of **linear methods**. We fit separate models within
treatment arms using only the treatment indicator and the linear predictor of
the internal risk prediction model. In the simpler case, we assume a constant
relative treatment effect (OR). Absolute benefit is then estimated from
$\text{expit}(lp +\log(\text{OR}))$, 
where $\text{expit}(x)=\frac{e^x}{1+e^x}$
and *lp* is the linear predictor of the prediction model. A different approach
fits a logistic regression using treatment, risk linear predictor and their
interaction within each treatment arm. In this case, absolute benefit is
estimated from
$\text{expit}(\beta_0+\beta_{lp}lp) - \text{expit}(\beta_0+\beta_{t_x}+(\beta_{lp}+\beta_*)lp)$.
We will refer to this method as the linear interaction approach.

@@

@@

The model including a constant relative treatment effect had the lowest median
RMSE in scenarios where the base case of true constant relative treatment effect
(OR = 0.8, N = 4250 and AUC = 0.75) or moderate relative deviations were
considered (Figure \@ref(fig:rmsebase); Panel A). However, when we considered
strong linear and quadratic deviations from the base case the linear interaction
model performed best (Figure \@ref(fig:rmsebase); Panels B and C). Only in the
case of strong quadratic deviations models including RCS smoothing (3 knots)
performed equally well to the linear interaction method. Increasing the number
of knots in RCS smoothing resulted in higher error rates across all scenarios.
The performance of the adaptive approach was quite comparable with the best
performing model in all scenarios.

@@

When we increased the sample size (N = 17000), in all scenarios the model more
in agreement with the underlying settings had the lowest error rates. Under the
base case of constant relative treatment effects (OR = 0.8), the model including
a constant relative treatment effect had the lowest RMSE (Figure
\@ref(fig:rmsesamplesize); Panel A). When introducing moderate and strong linear
deviations the linear interaction model performed best (Figure
\@ref(fig:rmsesamplesize); Panels B and C). In the case of true strong quadratic
deviations from the base case of constant relative effects, the more flexible
RCS smoothing models (3 knots) had the lowest RMSE. Again, the increased
flexibility of RCS smoothing with higher number of knots resulted in overfitting
and worse performance (Figure \@ref(fig:rmsesamplesize); Panel D).

@@

When we increased the true prediction AUC to 0.85, models including RCS
smoothing had the lowest RMSE when strong quadratic deviations from the base
case of true constant relative treatment effects were assumed (Figure
\@ref(fig:rmseauc); Panel D). However, with milder deviations, the linear
interaction model had the lowest RMSE with the RCS smoothing methods (3 knots)
being a close second (Figure \@ref(fig:rmseauc); Panels B and C). Increasing the
number of knots of RCS smoothing resulted in increased RMSE, which was less
pronounced in the case of strong quadratic deviations.

@@

All candidate methods demonstrated comparable discrimination for benefit in all
scenarios where linear and quadratic deviations from the base case of constant
treatment effect were considered (Figure \@ref(fig:discrimination)). However,
models including a linear interaction with the risk linear predictor tended to
present much lower variability compared to all other model-based and smoothing
approaches. We also observed an increasing trend of discrimination for benefit
variability with increasing number of restricted cubic spline knots in all
scenarios. This is evidence that the increased flexibility of these methods
often led to overfitting.

@@

In line with (Califf 1997; Steyerberg 2000), we fitted a logistic regression
model with 6 baseline covariates, i.e. age, Killip class, systolic blood
pressure, heart rate, an indicator of previous MI, and the location of MI, to
predict 30-day mortality risk. A constant effect of treatment was included in
the model. When deriving risk predictions for individuals we set the treatment
indicator to 0. More information on model development can be found in the
supplement.

>>>>

---
title: "Smooth risk-based predictive approaches to treatment effect heterogeneity: A simulation study"
output:
    bookdown::pdf_document2: default
    bookdown::word_document2:
        reference_docx: reference.docx
geometry: margin=1.0in
toc: false
font-size: 11pt
header-includes:
  - \renewcommand*\familydefault{\sfdefault}
  - \usepackage{setspace}
  - \usepackage{amsmath}
  - \doublespacing
  - \usepackage[left]{lineno}
  - \usepackage{amssymb}
  - \usepackage{bm}
  - \usepackage{booktabs}
  - \newcommand\given[1][]{\:#1\vert\:}
editor_options: 
  chunk_output_type: console
bibliography: references.bib
csl: jce.csl
---

@@

In the simulated datasets based on the base-case scenario treatment was
allocated at random using a 50/50 split. For each patient we simulated $8$
baseline covariates, where $x_1,\dots,x_4\sim N(0, 1)$ and
$x_5,\dots,x_8\sim B(1, 0.2)$. Outcomes for patients in the control arm were
generated from a logistic regression model including all baseline covariates.
Coefficient values were such, so that the prediction model had an AUC of $0.75$
and an event rate of $20%$ in the control arm was achieved. Outcomes in the
treatment arm were created using the same logistic regression model, including a
constant treatment effect odds ratio (OR) of $0.8$. The generated samples of the
base-case scenario were of size $n=4,250$ ($80%$ power for the detection of an
unadjusted OR of $0.8$).

@@

A true logistic regression model with a constant treatment effect (constant OR)
implies that outcome risk in the treatment arm is a straight line parallel to
the first diagonal on the *log-odds* scale, with distance equal to $\log(\text{OR})$. 
(See Figure XX of the supplementary material). We assessed the effect of
stronger and absent relative treatment effects ($\text{OR}=0.5$ or
$\text{OR}=1$). We also relaxed the assumption that the line should be parallel
to the diagonal, considering moderate and stronger linear deviations. Finally,
we dropped the assumption of linearity allowing for quadratic deviations.

@@

The **stratified HTE method** was suggested as an alternative to traditional
subgroup analyses. Patients are stratified into equally-sized risk strata---in
this case based on risk quartiles. Absolute effects are estimated using the
differences in event rates between treatments within risk quarters. We
considered this approach as a reference, expecting it to perform worse than the
other candidates, as its objective is not individualized benefit prediction.

We also considered a set of **linear methods**. We fit separate models within
treatment arms using only the treatment indicator and the linear predictor of
the internal risk prediction model. In the simpler case, we assume a constant
relative treatment effect (OR). Absolute benefit is then estimated from
$\text{expit}(lp +\log(\text{OR}))$, 
where $\text{expit}(x)=\frac{e^x}{1+e^x}$
and *lp* is the linear predictor of the prediction model. A different approach
fits a logistic regression using treatment, risk linear predictor and their
interaction. In this case, absolute benefit is estimated from
$\text{expit}(\beta_0+\beta_{lp}lp) - \text{expit}(\beta_0+\beta_{t_x}+(\beta_{lp}+\beta_*)lp)$.
We will refer to this method as the linear interaction approach.

@@

Assuming that $\tau(\bm{x})=E\{y\given\bm{x}, t_x=0\} - E\{y\given\bm{x},t_x=1\}$ is the true benefit for each patient and $\hat{\tau}(x)$ is the estimated benefit from a method under study, the ideal loss function to use for the considered methods would be the unobservable root mean squared error $E\big\{(\hat{\tau} - \tau)^2\given \bm{x}\big\}$. However, in simulations the true patient-level benefit is available. Therefore, we estimated RMSE from
$$\text{RMSE}=\frac{1}{n}\sum_{i=1}^n\big(\tau(\bm{x}_i) - \hat{\tau}(\bm{x}_i)\big)^2$$

We also compared the predictive performance of the methods under study. We
assessed discrimination using the c-for-benefit statistic described in
[@vanKlaveren2018]. Patients in each treatment arm are ranked based on their
predicted benefit and then are matched 1:1, dropping patients in the larger
treatment arm without a pair. We define *observed* pair-specific treatment
benefit as the difference of observed outcomes between the untreated and the
treated patient of each pair. Pair-specific *predicted* benefit is defined as
the average of predicted benefits within each pair. Then, c-for-benefit is
defined as the probability that from two randomly chosen matched patient pairs
with unequal observed benefit, the pair with greater observed benefit also has a
higher predicted benefit.

We evaluated calibration in a similar manner, using the integrated calibration
index (ICI) for benefit [@Austin2019]. After creating pairs based on predicted
benefit, observed benefits are regressed on the predicted benefits using a
locally weighted scatterplot smoother (loess). The ICI is the area between the
loess fit and the diagonal, with values closer to $0$ indicating better
calibration.

@@

The model including a constant relative treatment effect had the lowest median
RMSE in scenarios where the base case of true constant relative treatment effect
(OR = 0.8, N = `r 4250` and AUC = 0.75) or moderate relative deviations were
considered (Figure \@ref(fig:rmsebase); Panel A). However, when we considered
strong linear and quadratic deviations from the base case the linear interaction
model performed best (Figure \@ref(fig:rmsebase); Panels B and C). Only in the
case of strong quadratic deviations models including RCS smoothing (3 knots)
performed equally well to the linear interaction method. Increasing the number
of knots in RCS smoothing resulted in higher error rates across all scenarios.
The performance of the adaptive approach was quite comparable with the best
performing model in each scenario.

@@

When we increased the sample size (N = `r 17000`), in all scenarios the model more
in agreement with the underlying settings had the lowest error rates. Under the
base case of constant relative treatment effects (OR = 0.8), the model including
a constant relative treatment effect had the lowest RMSE (Figure
\@ref(fig:rmsesamplesize); Panel A). When introducing moderate and strong linear
deviations the linear interaction model performed best (Figure
\@ref(fig:rmsesamplesize); Panels B and C). In the case of true strong quadratic
deviations from the base case of constant relative effects, the more flexible
RCS smoothing models (3 knots) had the lowest RMSE. Again, the increased
flexibility of RCS smoothing with higher number of knots resulted in overfitting
and worse performance (Figure \@ref(fig:rmsesamplesize); Panel D).

@@

When we increased the true prediction AUC to 0.85, models including RCS
smoothing had the lowest RMSE in the presence of strong quadratic deviations
from the base case of true constant relative treatment effects (Figure
\@ref(fig:rmseauc); Panel D). However, with milder deviations, the linear
interaction model had the lowest RMSE with the RCS smoothing methods (3 knots)
being a close second (Figure \@ref(fig:rmseauc); Panels B and C). Increasing the
number of knots of RCS smoothing resulted in increased RMSE, which was less
pronounced in the case of strong quadratic deviations.

@@

All candidate methods demonstrated comparable discrimination for benefit in all
scenarios where linear and quadratic deviations from the base case of constant
treatment effect were considered (N=`r 4250`; AUC = `r 0.75`). However, models
including a linear interaction with the risk linear predictor tended to present
much lower variability compared to all other model-based and smoothing
approaches (Figure \@ref(fig:discrimination)). We also observed an increasing
trend of discrimination for benefit variability with increasing number of
restricted cubic spline knots in all scenarios.

@@

In line with previous analyses [@Califf1997; @Steyerberg2000], we fitted a logistic regression
model with 6 baseline covariates, i.e. age, Killip class, systolic blood
pressure, heart rate, an indicator of previous MI, and the location of MI, to
predict 30-day mortality risk. A constant effect of treatment was included in
the model. When deriving risk predictions for individuals we set the treatment
indicator to 0. More information on model development can be found in the
supplement.
